---
title: "Warmup 2: Basic Manipulation of Data Tables"
author: "Felicia Liu"
date: "2/13/2021"
output: html_document
---

# 1) Data "storms"

```{r}
library(dplyr)
library(ggplot2)
```

a)
```{r}
num_rows <- dim(storms)[1]
num_rows
```

b)
```{r}
num_cols <- dim(storms)[2]
num_cols
```

c)
```{r}
storms 
```
In R, a tibble is a data frame that are a little simpler to use compared to original data frames. Tibbles usually display 10 rows (if there are over 10 rows total) and they do not change an input's type or adjust the name of variables. The arguments are also evaluated sequentially and only vectors of length 1 are recycled since recycling vectors with greater lengths likely causes bugs.

d) These abbreviations between angle brackets tell us what type of data the column contains. For example, the lat column shows dbl in brackets underneath, which means that the values in the column are all doubles. This means that they are numbers that may contain a floating point value (a decimal). On the other hand, the column status shows chr in brackets underneath, and this means that the values in the column status are all characters, which is essentially a string of 1 or more characters. 


# 2) Storms in 1980

a)
```{r}
num_storms_1980 <- dim(filter(storms, year == '1980'))[1]
num_storms_1980
```
b)
```{r}
unique_names_1980 <- distinct(filter(storms, year == '1980'), name)
unique_names_1980
```
c)
```{r}
sorted_names_1980 <- unique_names_1980 %>% arrange(desc(name))
sorted_names_1980
```
d)
```{r}
storm_records_1980 <- filter(storms, year == '1980') %>% 
  group_by(name) %>% summarize(count = n()) %>% arrange(count)
storm_records_1980
```

# 3) More Manipulation

a)
```{r}
storm_names_1990s <- filter(storms, year >= 1990 & year <= 1999) %>%
  group_by(name, year) %>% summarize()
head(storm_names_1990s, 5)
tail(storm_names_1990s, 5)
```


b)
```{r}
storms_per_year <- storms %>% group_by(year) %>% summarize(num_storms = n())
head(storms_per_year, 15)
```


c)
```{r}
storm_records_per_year <- storms %>% group_by(name, year) %>%
  summarize(count = n())
head(storm_records_per_year, 10)
tail(storm_records_per_year, 10)
```


# 4) Some basic graphs with "ggplot2"

a) 
```{r}
storms_1990s <- filter(storms, year >= 1990 & year <= 1999)
ggplot(storms_1990s, aes(x = as.character(year))) + geom_bar() +
  labs(y = "Number of Storms", x = "Year") 
```

b)
```{r}
ggplot(storms, aes(x = wind)) + geom_histogram(color = "red",
  fill = "yellow", binwidth = 5) +
  labs(x = "Storm's maximum sustained wind speed (in knots)",
       y = "Number of storms")
```

c)
```{r}
ggplot(storms, aes(x = pressure)) + geom_density(fill = "pink") +
  labs(x = "Air pressure at the storm's center (in millibars)")
```

d)
```{r}
storms_1990_2009 <- filter(storms, year >= 1990 & year <= 2009)
ggplot(storms_1990_2009, aes(x = pressure)) + geom_density() +
  facet_wrap(~ year)
```

e)
```{r}
avg_wind_by_year <- storms %>% group_by(year) %>% 
  summarize(avg_wind_speed = mean(wind))
ggplot(avg_wind_by_year, aes(x = year, y = avg_wind_speed)) + 
  geom_point() + geom_line() + 
  labs(x = "Year", y = "Average Wind-Speed (in knots)")
```


# 5) Storm Status and Categories

a)
```{r}
storm_status <- storms %>% group_by(status) %>% summarize() 
storm_status
```


b)
```{r}
storm_categories <- storms %>% group_by(category) %>% summarize() 
storm_categories
```


c)
```{r}
storm_combos <- storms %>% group_by(status, category) %>% summarize() 
storm_combos
```


d)
```{r}
storms_4_cols <- storms %>% group_by(category, status) %>% 
  summarize(avg_pressure= mean(pressure), avg_wind = mean(wind))
storms_4_cols
```


e)
```{r}
ggplot(storms, aes(x = status, y = pressure) ) + geom_boxplot()
```



# 6) Wind Speed and Pressure
```{r}
ggplot(storms, aes(x = wind, y = pressure)) + 
  geom_point(aes(col = category))
```


# 7) Maximum Wind speeds
```{r}
max_wind_per_storm <- storms %>% group_by(name, year) %>% 
  summarize(max_wind = max(wind))
head(max_wind_per_storm, 10)
tail(max_wind_per_storm, 10)
```


# 8) Maximum Wind Speed in each year
```{r}
max_wind_per_year <- storms %>% group_by(year, name) %>%
  summarize(max_wind = max(wind)) %>% 
  filter(max_wind == max(max_wind), .by_group = TRUE) %>%
  arrange(desc(max_wind))
head(max_wind_per_year, 10)
tail(max_wind_per_year, 10)
```

