---
title: 'Lab 1: Getting started with R and RStudio'
subtitle: "Stat 133, Fall 2020"
output: html_document
urlcolor: blue
---

> ### Learning Objectives:
>
> - Get started with R as a scientific calculator
> - Understand pane layout of RStudio
> - Understand the help documentation in R
> - How to install packages
> - Difference between `.R` and `.Rmd` files
> - Get to know markdown syntax

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## 1) R and RStudio

- You should have installed __R__
	- R for Mac: [https://cran.r-project.org/bin/macosx/](https://cran.r-project.org/bin/macosx/)
	- R for windows: [https://cran.r-project.org/bin/windows/base/](https://cran.r-project.org/bin/windows/base/)

- You should also have installed __RStudio__
	- RStudio download: [https://www.rstudio.com/products/rstudio/download/](https://www.rstudio.com/products/rstudio/download/)

- R and RStudio are two different things
	- R is both a software and language (like python, matlab, julia, etc)
	- R comes with its minimal Graphical User Interface (R-GUI), but we won't use this.
	- Instead, we are going to use RStudio as our main working environment. 
	- RStudio is much more than a simple GUI; RStudio is technically an _Integrated Development Environment_ (IDE).
	- We are going to use mainly RStudio, which provides a nice working environment and development framework.



## 2) R as a scientific calculator

Launch RStudio and notice the default position of the panes (or panels):

- Console (entire left)
- Environment/History (tabbed in upper right)
- Files/Plots/Packages/Help (tabbed in lower right)

__FYI:__ you can change the default location of the panes, 
among many other things: [Customizing RStudio](https://support.rstudio.com/hc/en-us/articles/200549016-Customizing-RStudio). If you have no experience working with R/RStudio, you don't have to customize anything right now. It's better if you wait some days until you get a better feeling of the working environment. You will probably be experimenting (trial and error) some time with the customizing options until you find what works for you.


### 2.1) First contact with the R console

If you have never used software in which you have to type commands and code, our best suggestion is that you begin typing basic things in the __console__, using R as a scientific calculator.

For instance, consider the monthly bills of Leia (a fictitious stats undergrad student): 

- cell phone $80
- transportation $20
- groceries $527
- gym $10
- rent $1500
- other $83

You can use R to find Leia's total expenses by typing these commands in the console:

```{r}
# total expenses
80 + 20 + 527 + 10 + 1500 + 83
```

Often, it will be more convenient to create __objects__ or __variables__ 
that store one or more values. To do this, type the name of the variable, 
followed by the assignment operator `<-`, followed by the assigned value.
For example, you can create an object `phone` for the cell phone bill, 
and then inspect the object by typing its name:

```{r}
# objects with assigned values
phone <- 80
phone
```

All R statements where you create objects are known as "assignments", and they have this form:

```r
object <- value
```

this means you assign a `value` to a given `object`;
you can read the previous assignment as "phone gets 80".

RStudio has a keyboard shortcut for the arrow operator `<-`: 
`Alt` + `-` (the minus sign).

Notice that RStudio automagically surrounds `<-` with spaces, 
which demonstrates a useful code formatting practice. So do yourself 
(and others) a favor by ALWAYS surrounding an assignment operator with spaces.

Alternatively, you can also use the equals sign `=` as an assignment 
operator:

```{r}
coffee = 30
coffee
```

You will be working with RStudio a lot, and you will have time to learn
most of the bells and whistles RStudio provides. Think about RStudio as
your "workbench". Keep in mind that RStudio is NOT R. RStudio is an environment
that makes it easier to work with R, while taking care of many of
the little tasks than can be a hassle.



### 2.2) Your Turn

- Make more assignments to create variables `transportation`, `groceries`, `gym`, 
`rent`, and `other` with their corresponding amounts.

```{r}
# create variables
transportation <- 20
groceries <- 527
gym <- 10
rent <- 1500
other <- 83
```

- Now that you have all the variables, create a `total` object with the 
sum of the expenses.

```{r}
# total expenses
total <- phone + transportation + groceries + gym +
  rent + other
total
```

- Assuming that Leia has the same expenses every month, how much would she 
spend during a school "semester"? (assume the semester involves five months).

```{r}
# semester expenses
semester <- total * 5
semester
```

- Maintaining the same assumption about the monthly expenses, how much would Leia
spend during a school "year"? (assume the academic year is 10 months).

```{r}
# year expenses
year <- total * 10
year
```



### 2.3) Object Names

There are certain rules you have to follow when creating objects and variables.
Object names cannot start with a digit and cannot contain certain other characters 
such as a comma or a space. You will be wise to adopt a convention for demarcating 
words in names.


```r
i_use_snake_case
other.people.use.periods
evenOthersUseCamelCase
```

The following are invalid names (and invalid assignments)

```
# cannot start with a number
5variable <- 5

# cannot start with an underscore
_invalid <- 10

# cannot contain comma
my,variable <- 3

# cannot contain spaces
my variable <- 1
```

This is fine but a little bit too much:

```r
this_is_a_really_long_name <- 3.5
```


### 2.4) Functions

R has many functions. To use a function type its name followed by parenthesis.
Inside the parenthesis you pass an input. Most functions will produce some
type of output:

```r
# absolute value
abs(10)
abs(-4)

# square root
sqrt(9)

# natural logarithm
log(2)
```


### 2.5) Comments in R

All programming languages use a set of characters to indicate that a
specifc part or lines of code are __comments__, that is, things that are
not to be executed. R uses the hash or pound symbol `#` to specify comments.
Any code to the right of `#` will not be executed by R.

```r
# this is a comment
# this is another comment
2 * 9

4 + 5  # you can place comments like this
```


### 2.6) Case Sensitive

R is case sensitive. This means that `phone` is not the same
as `Phone` or `PHONE`

```{r}
# case sensitive
phone <- 80
Phone <- -80
PHONE <- 8000

phone + Phone
PHONE - phone
```


### 2.7) Your turn

Take your objects (i.e. variables) `phone`, `transportation`, `groceries`, 
`gym`, `rent`, and `other` and pass them inside the _combine_ function `c()` 
to create a vector `expenses`.

```{r}
# vector expenses
expenses <- c(phone, transportation, groceries, gym, rent, other)
```

Now, use the graphing function `barplot()` to produce a barchart of 
`expenses`:

```{r eval = FALSE}
barplot(expenses)
```

You can see the manual (help) document of `barplot()` by typing `?barplot`;
look at the arguments of this function and learn how to:

use `col` to color the bars

```{r}
# barchart with colored bars
barplot(expenses, col = 2) 
```

use , `horiz` to get a barchart in horizontal orientation

```{r}
# barchart horizontally oriented
barplot(expenses,  col = 2, horiz = TRUE)
```

use `names.arg` to add labels to the bars (with names of the variables below 
each of the bars)

```{r}
# barchart with bar labels
barplot(expenses, col = 2, horiz = TRUE, 
        names.arg = c('phone', 'transportation', 'groceries', 'gym', 
                      'rent', 'other'))
```


Find out how to use `sort()` to sort the elements in `expenses`, in order to
produce a bar-chart with bars in decreasing order. Also 
optional, see if you can find out how to display the values of each variable
at the top of each bar.

```{r}
# barchart in decreasing/increasing order
barplot(sort(expenses, decreasing = FALSE),col = 2, horiz = TRUE,
        names.arg = c('gym', 'transportation', 'phone', 'other', 
                      'groceries', 'rent'))
```




## 3) Introduction to R Markdown files

Most of the times you won't be working directly on the console.
Instead, you will be typing your commands in some __source file__.
The most basic type of source files are known as _R script files_. 
But there are more flavors of source files. A very convenient type of source
file that allow you to mix R code with narrative is an __R markdown file__
commonly referred to as `Rmd` file.


### 3.1) Get to know the `Rmd` files

In the menu bar of RStudio, click on __File__, then __New File__,
and choose __R Markdown__. Select the default option (Document),
and click __Ok__.

__Rmd__ files are a special type of file, referred to as a _dynamic document_,
that allows to combine narrative (text) with R code. Because you will
be turning in most homework assignments as `Rmd` files, it is important
that you quickly become familiar with this resource.

Locate the button __Knit HTML__ (the one with a knitting icon) and click on it
so you can see how `Rmd` files are renderer and displayed as HTML documents.

_R markdown_ files use a special syntax called __markdown__. To be more precise, 
Rmd files let you type text using either: 1) R syntax for code that needs to be 
executed; 2) markdown syntax to write your _narrative_, and 3) latex syntax for 
math equations and symbols.

You will have time to learn the basics of this syntax during the semester, and
we expect that feel comfortable with markdown at the end of the course.



### 3.2) Your turn

Open a new `Rmd` file in the _source_ pane,and include all the previous commands 
in separated code chunks. Knit the file to get an html document. These are the 
files that you have to submit to bCourses when you finish the lab work.



## 4) Installing Packages

R comes with a large set of functions and packages. A package is a collection
of functions that have been designed for a specific purpose. One of the great
advantages of R is that many analysts, scientists, programmers, and users
can create their own pacakages and make them available for everybody to use them.
R packages can be shared in different ways. The most common way to share a
package is to submit it to what is known as __CRAN__, the
_Comprehensive R Archive Network_.

You can install a package using the `install.packages()` function. To do this, 
we recommend that you run this command directly on the console. Do NOT include 
this command in a code chunk of an Rmd file: you will very likely get an error
message when knitting the Rmd file.

To use `install.packages()` just give it the name of a package, surrounded by qoutes, 
and R will look for it in CRAN, and if it finds it, R will download it to your computer.

```r
# installing (run this on the console!)
install.packages("knitr")
```

You can also install a bunch of packages at once:

```r
# run this command on the console!
install.packages(c("readr", "ggplot2"))
```

Once you installed a package, you can start using its functions
by _loading_ the package with the function `library()`. By the way, when 
working on an Rmd file that uses functions from a given package, you MUST include 
a code chunk with the `library()` command.

```r
# (this command can be included in an Rmd file)
library(knitr)
```
```{r}
library(knitr)
```



## 5) Getting Help

Because we work with functions all the time, it's important to know certain
details about how to use them, what input(s) is required, and what is the
returned output.

There are several ways to get help.

If you know the name of a function you are interested in knowing more,
you can use the function `help()` and pass it the name of the function you
are looking for:

```r
# documentation about the 'abs' function
help(abs)

# documentation about the 'mean' function
help(mean)
```

Alternatively, you can use a shortcut using the question mark `?` followed
by the name of the function:

```r
# documentation about the 'abs' function
?abs

# documentation about the 'mean' function
?mean
```

- How to read the manual documentation
	- Title
	- Description
	- Usage of function
	- Arguments
	- Details
	- See Also
	- Examples!!!

`help()` only works if you know the name of the function your are looking for.
Sometimes, however, you don't know the name but you may know some keywords.
To look for related functions associated to a keyword, use double
`help.search()` or simply `??`

```r
# search for 'absolute'
help.search("absolute")

# alternatively you can also search like this:
??absolute
```
Notice the use of quotes surrounding the input name inside `help.search()`



### 5.1) Your Turn: Pythagoras formula

The pythagoras formula is used to compute the length of the hypotenuse, $c$, of 
a right triangle with legs of length $a$ and $b$.

$$
c = \sqrt{a^2 + b^2}
$$

Calculate the hypotenuse of a right triangle with legs of length 3 and 4.
Use the `sqrt()` function, and create variables `a = 3` and `b = 4`.
If you don't know what's the symbol to calculate exponents, search for the 
help documentation of the arithmetic operators: `?Arithmetic`.

```{r}
a <- 3
b <- 4
c <- sqrt(3**2 + 4**2)
```


### 5.2) Your Turn: Binomial Formula

The formula for the binomial probability is:

$$
Pr(k; n, p) = Pr(X = k) = {n \choose k} p^k (1-p)^{n-k}
$$

where:

- $n$ is the number of (fixed) trials
- $p$ is the probability of success on each trial
- $1 - p$ is the probability of failure on each trial
- $k$ is a variable that represents the number of successes out of $n$ trials
- the first term in parenthesis is not a fraction, it is the number of combinations in which $k$ success can occur in $n$ trials

R provides the `choose()` function to compute the number of combinations: 

$$
{n \choose k} = \frac{n (n-1) \cdots (n - k +1)}{k (k-1) \cdots 1}
$$

For instance, the number of combinations in which $k$ = 2 
success can occur in $n$ = 5 trials is:

```{r}
choose(n = 5, k = 2)
```

Combinations are typically expressed in terms of factorials as:

$$
\frac{n!}{k! (n - k)!}
$$

Conveniently, R also provides the function `factorial()` to calculate the factorial of an integer:

```{r}
factorial(4)
```

Let's consider a simple example. A fair coin is tossed 5 times. What is the probability 
of getting exactly 2 heads?

- Create the objects `n`, `k`, and `p` for the number of trials, the number of 
success, and the probability of success, respectively.

- Use `factorial()` to compute the number of combinations "_n choose k_"

- Apply the binomial formula, using `factorial()`, to calculate the probability 
of getting exactly 2 heads out of 5 tosses.

- Recalculate the same probability but now using `choose()` 
(instead of `factorial()`)

- Consider rolling a fair die 10 times. What is the probability 
of getting exactly 3 sixes?

- Now look for help documentation (e.g. `help.search()` or `??`) using the 
keyword binomial: `binomial`.

- You should get a list of topics related with the searched term `binomial`.

- Choose the one related with the _Binomial Distribution_, which is 
part of the R package `stats` (i.e. `stats::Binomial`).

- Read the documentation and figure out how to use the `dbinom()` function 
to obtain the above probabilities: 2 heads in 5 coin tosses, and 
3 sixes in 3 rolls of a die.

- How would you modify the previous binomial function to calculate the 
same probability (2 heads in 5 tosses) of a __biased__ coin with a chance of 
heads of 35%?

- Finally, obtain the probability of getting more than 3 heads in 5 tosses
with a biased coin of 35% chance of heads.

```{r} 
#A fair coin is tossed 5 times. What is the probability of getting exactly 
#2 heads?
n <- 5
k <- 2
p <- 0.5
combinations <- factorial(n)/(factorial(k)*factorial(n-k))
probability <- combinations*(p**k)*(1-p)**(n-k)
probability_2 <- choose(n, k)*(p**k)*(1-p)**(n-k)
```
```{r}
dbinom(x = 2, size = 5, prob = 0.5)
```


```{r}
#Consider rolling a fair die 10 times. 
#What is the probability of getting exactly 3 sixes?
dbinom(x = 3, size = 10, prob = 1/6)
```
```{r}
#Probability of 2 heads in 5 tosses of a biased coin
#with a chance of heads of 35%
dbinom(x = 2, size = 5, prob = 0.35)
```
```{r}
#Probability of getting more than 3 heads in 5 tosses
#with a biased coin of 35% chance of heads
1 - pbinom(q = 2, size = 5, prob = 0.35)
```


### 5.3) Your turn

This part doesn't need to be included in your Rmd file. Instead, type commands 
directly on the console:

- On the console, install packages `"stringr"`, `"RColorBrewer"`, and "`xml2`"

- Calculate: $3x^2 + 4x + 8$ when $x = 2$

- Calculate: $3x^2 + 4x + 8$ but now with a numeric sequence for $x$ 
using `x <- -3:3`

- Find out how to look for information about math binary operators
like `+` or `^` (without using `?Arithmetic`).

- There are several tabs in the pane `Files, Plots, Packages, Help, Viewer`.
Find what does the tab __Files__ is good for?

- What about the tab __Help__? 

- In the tab __Help__, what happens when you click the button with a House icon?

- Now go to the tab __History__. What is it good for? and what about the 
buttons of its associated menu bar?

- Likewise, what can you say about the tab __Environment__?


