---
title: "Lab 2: First contact with dplyr and ggplot2"
subtitle: "Stat 133, Fall 2020"
output: html_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```

> ### Learning Objectives:
>
> - Get started with `"dplyr"` and its basic verbs:
>   + `slice()`, `filter()`, `select()`
>   + `mutate()`
>   + `arrange()`
>   + `summarise()`
>   + `group_by()`
> - Get started with `"ggplot2"`
> - Produce basic plots with `ggplot()`


### General Instructions

- Write your descriptions, explanations, and code in an `Rmd` (R markdown) file.
- Name this file as `lab02-first-last.Rmd`, where `first` and `last` are your
first and last names (e.g. `lab02-gaston-sanchez.Rmd`).
- Knit your `Rmd` file as an html document (default option).
- Submit your `Rmd` and `html` files to bCourses, in the corresponding lab assignment.




## 1) Manipulating and Visualizing Data Frames

In this lab, you will start learning a couple of approaches to manipulate tables
and create statistical graphics. We are going to use the functionality of the 
package `"dplyr"` to work with tabular data (_in a syntactic way_). This is a 
fairly recent package introduced a couple of years ago, but it is based on more
than a decade of research and work lead by [Hadley Wickham](http://hadley.nz/).

Likewise, to create graphics in a fairly consistent and visually pleasing way, 
we are going to use the package `"ggplot2"`, also originally authored by 
Hadley Wickham, and developed as part of his PhD more than a decade ago.

While you follow this lab, you may want to open these cheat sheets:

- [dplyr cheatsheet](https://github.com/ucb-stat133/stat133-cheatsheets/blob/master/data-transformation-cheatsheet.pdf)
- [ggplot2 cheatsheet](https://github.com/ucb-stat133/stat133-cheatsheets/blob/master/ggplot2-cheatsheet-2.1.pdf)




### 1.1) Installing packages

I'm assuming that you already installed the packages `"dplyr"` and `"ggplot2"`. 
If that's not the case then __run on the console__ the command below 
(do NOT include this command in your `Rmd`):

```r
# don't include this command in your Rmd file
# don't worry too much if you get a warning message
install.packages(c("dplyr", "ggplot2"))
```

Remember that you only need to install a package once! After a package has been 
installed in your machine, there is no need to call `install.packages()` again 
on the same package. What you should always invoke, in order to use the 
functions in a package, is the `library()` function:

```{r eval=FALSE}
# (include these commands in your Rmd file)
# load the packages
library(dplyr)
library(ggplot2)
```

__About loading packages:__ Another rule to keep in mind is to always load any 
required packages at the very top of your script files (`.R` or `.Rmd` or `.Rnw` files). 
Avoid calling the `library()` function in the middle of a script. Instead, 
load all the packages before anything else.



### 1.2) Data `starwars`

The data file for this lab has to do with Star Wars characters. 
The dataset is part of the `dplyr` package: `starwars`. So, assuming that you
loaded the package `"dplyr"`, then simply type the name of the object:
`starwars`

```{r read, eval = FALSE}
# assuming you loaded dplyr ...
starwars
```



# Part I) Basic `"dplyr"` verbs

To make the learning process of `"dplyr"` gentler, Hadley Wickham proposes 
beginning with a set of five _basic verbs_ or operations for data frames 
(each verb corresponds to a function in `"dplyr"`):

- __filter__: keep rows matching criteria
- __select__: pick columns by name
- __mutate__: add new variables
- __arrange__: reorder rows
- __summarise__: reduce variables to values

I've slightly modified Hadley's list of verbs: 

- `filter()`, `slice()`, and `select()`: subsetting and selecting rows and columns
- `mutate()`: add new variables
- `arrange()`: reorder rows
- `summarise()`: reduce variables to values
- `group_by()`: grouped (aggregated) operations



## 2) Filtering, slicing, and selecting

`slice()` allows you to select rows by position:

```{r}
# first three rows
three_rows <- slice(starwars, 1:3)
three_rows
```

`filter()` allows you to select rows by defining a condition (which could be 
simple or compound):

```{r}
# subset rows given a simple condition
# (height greater than 200 cm)
gt_200 <- filter(starwars, height > 200)
gt_200
```

```{r}
# subset rows given a compound condition
filter(starwars, height > 200 & mass < 100)
```


`select()` allows you to select one or more columns by name:

```{r}
# columns by name
name_height <- select(starwars, name, height)
```


### 2.1) Your turn:

- use `slice()` to subset the data by selecting the first 5 rows.

```{r}
# your code
five_rows <- slice(starwars, 1:5)
five_rows
```

- use `slice()` to subset the data by selecting rows 10, 15, 20, ..., 50. 
_Optional hint_: `seq()` is your friend.

```{r}
# your code
by_five <- slice(starwars, seq(10, 50, by = 5))
by_five
```

- use `slice()` to subset the data by selecting the last 5 rows.

```{r}
# your code
last_five <- slice(starwars, tail(row_number(), 5))
last_five
```

- use `filter()` to subset those individuals with height less than 100 cm tall.

```{r}
# your code
lt_100 <- filter(starwars, height > 100)
lt_100
```

- use `filter()` to subset rows of female individuals (`gender`).

```{r}
# your code
female <- filter(starwars, gender == 'feminine')
female
```

- use `filter()` to subset rows of individuals with brown hair color.

```{r}
# your code
brown_hair <- filter(starwars, hair_color == 'brown')
brown_hair
```

- use `filter()` and then `select()`, to subset rows of individuals from `Naboo`,
and then display their names.

```{r}
# your code
names_from_naboo = filter(starwars, homeworld == 'Naboo') %>% select('name')
names_from_naboo
```

- use `"dplyr"` functions to display the names of individuals with green
skin color.

```{r}
# your code
green_skin <- filter(starwars, skin_color == 'green') %>% select('name')
green_skin
```

- find how to `select()` the name, height, and mass, of male individuals, 
with brown or black hair color.

```{r}
# your code
selected_criteria <- filter(starwars, sex == 'male' & (hair_color == 'brown' | 
                            hair_color == 'black')) %>% 
  select(c('name', 'height', 'mass'))
selected_criteria
```

- find how to select the name and homeworld, of human female individuals.

```{r}
# your code
human_female_hmwrld <- filter(starwars, species == 'Human' & sex == 'female') %>%
  select(c('name', 'homeworld'))
human_female_hmwrld
```



## 3) Adding new variables: `mutate()`

Another basic verb is `mutate()` which allows you to add new variables.
Let's create a small data frame for the female individuals with three columns: 
`name`, `height`, and `mass`:

```{r}
# creating a small data frame step by step
fem <- filter(starwars, sex == 'female')
fem <- select(fem, name, height, mass)
fem <- slice(fem, c(1, 2, 5, 6, 8))
fem
```

Now, let's use `mutate()` to (temporarily) add a column with the ratio 
`height / mass`:

```{r results='hide'}
mutate(fem, height / mass)
```

You can also give a new name, like: `ht_wt = height / mass`:

```{r results='hide'}
mutate(fem, ht_wt = height / mass)
```

In order to permanently change the data, you need to assign the changes to 
an object:

```{r results='hide'}
fem2 <- mutate(fem, ht_m = height * 0.0254, wt_kg = mass * 0.4536)
fem2
```



## 4) Reordering rows: `arrange()`

The next basic verb of `"dplyr"` is `arrange()` which allows you to reorder 
rows. For example, here's how to arrange the rows of `fem` by `height`

```{r results='hide'}
# order rows by height (increasingly)
arrange(fem, height)
```

By default `arrange()` sorts rows in increasing order. To arrange rows in 
descending order you need to use the auxiliary function `desc()`.

```{r results='hide'}
# order rows by height (decreasingly)
arrange(fem, desc(height))
```

```{r results='hide'}
# order rows by height, and then mass
arrange(fem, height, mass)
```



### 4.1) Your Turn:

- using the data frame `fem`, add a new variable `product` with the product of 
`height` and `mass`.

```{r}
# your code
fem_with_product <- mutate(fem, product = height * mass)
fem_with_product
```

- create a new data frame `fem3`, by adding columns `log_height` and `log_mass`
with the log transformations of `height` and `mass`.

```{r}
# your code
fem3 <- mutate(fem, log_height = log(height), log_mass = log(mass))
fem3
```

- use the original data frame to `filter()` and `arrange()` those individuals 
with height less than 150 cm tall, in increasing order by height.

```{r}
# your code
starwars_heights_ordered <- filter(starwars, height < 150) %>% arrange(height)
starwars_heights_ordered
```

- display the name, homeworld, and species, of the top-5 tallest individuals.

```{r}
# your code
top_5_height <- arrange(starwars, desc(height)) %>% slice(1:5) %>% 
  select(c('name', 'homeworld', 'species'))
top_5_height
```

- display the name, homeworld, and species, for the top-5 heaviest individuals.

```{r}
# your code
top_5_mass <- arrange(starwars, desc(mass)) %>% slice(1:5) %>% 
  select(c('name', 'homeworld', 'species'))
top_5_mass
```




## 5) Summarizing values with `summarise()`

The next verb is `summarise()`. Conceptually, this involves applying a function
on one or more columns, in order to summarize values. This is probably easier 
to understand with one example.

Say you are interested in calculating the average height of all individuals.
To do this "a la dplyr" you use `summarise()`, or its synonym function 
`summarize()`:

```{r}
# average height (removing missing values)
summarise(starwars, avg_height = mean(height, na.rm = TRUE))
```

What if you want to calculate some 
summary statistics for `height`: min, median, mean, and max?

```{r}
# some stats for height (dplyr)
summarise(
  starwars, 
  min = min(height, na.rm = TRUE),
  median = median(height, na.rm = TRUE),
  avg = mean(height, na.rm = TRUE),
  max = max(height, na.rm = TRUE)
)
```



## 6) Grouped operations

To actually appreciate the power of `summarise()`, we need to introduce the 
other major basic verb in `"dplyr"`: `group_by()`. This is the function that 
allows you to perform data aggregations, or _grouped operations_.

Let's see the combination of `summarise()` and `group_by()` to calculate the
average salary by team:

```{r results='hide'}
# average height, grouped by homeworld
summarise(
  group_by(starwars, homeworld),
  avg_salary = mean(height, na.rm = TRUE)
)
```


Here's a more fancy example: average mass and height, by homeworld, displayed
in desceding order by average height:

```{r results='hide'}
arrange(
  summarise(
    group_by(starwars, homeworld),
    avg_height = mean(height, na.rm = TRUE),
    avg_mass = mean(mass, na.rm = TRUE)),
  desc(avg_height)
)
```


### 6.1) Your turn:

- use `summarise()` to get the largest height value.

```{r}
# your code
summarise(starwars, largest_height = max(height, na.rm = TRUE))
```

- use `summarise()` to get the standard deviation of `mass`.

```{r}
# your code
summarise(starwars, sd = sd(mass, na.rm = TRUE))
```

- use `summarise()` and `group_by()` to display the median of mass,
by homeworld

```{r}
# your code
summarise(group_by(starwars, homeworld),
                   median_mass = median(mass, na.rm = TRUE))
```

- display the average mass by gender, in ascending order,

```{r}
# your code
arrange(summarise(group_by(starwars, gender),
                   avg_mass = mean(mass, na.rm = TRUE)), avg_mass)
```

- obtain the mean and standard deviation of `height`, for female characters.

```{r}
# your code
summarise(filter(starwars, sex == 'female'), avg_height =
            mean(height, na.rm = TRUE), sd_height = sd(height, na.rm = TRUE))
```



# Part II) First contact with `ggplot()`

The package `"ggplot2"` is probably the most popular package in R to create
_beautiful_ static graphics. Compared to the functions in the base package 
`"graphics"`, the package `"ggplot2`" follows a somewhat different philosophy, 
and it tries to be more consistent and modular as possible.

- The main function in `"ggplot2"` is `ggplot()`

- The main input to `ggplot()` is a data frame object.

- You can use the internal function `aes()` to specify what columns of the data 
frame will be used for the graphical elements of the plot.

- You must specify what kind of _geometric objects_ or __geoms__ will be
displayed: e.g. `geom_point()`, `geom_bar()`, `geom_boxpot()`.

- Pretty much anything else that you want to add to your plot is controlled 
by auxiliary functions, especially those things that have to do with the format,
rather than the underlying data.

- The construction of a ggplot is done by _adding layers_ with the `+` operator.



## 7) Scatterplots

Let's start with a scatterplot of `height` and `mass`

```{r scatter1}
# scatterplot (option 1)
ggplot(data = starwars) +
  geom_point(aes(x = height, y = mass))
```

- `ggplot()` creates an object of class `"ggplot"`
- the main input for `ggplot()` is `data` which must be a data frame
- then we use the `"+"` operator to add a layer
- the geometric object (geom) are points: `geom_points()`
- `aes()` is used to specify the `x` and `y` coordinates, by taking columns
`points` and `salary` from the data frame

The same scatterplot can also be created with this alternative, and more 
common use of `ggplot()`

```{r eval = FALSE}
# scatterplot (option 2)
ggplot(data = starwars, aes(x = height, y = mass)) +
  geom_point()
```


### 7.1) Adding color

Say you want to color code the points in terms of `gender`

```{r scatter2}
# colored scatterplot 
ggplot(data = starwars, aes(x = height, y = mass)) +
  geom_point(aes(color = gender))
```


### 7.2) Your turn:

- Open the ggplot2 cheatsheet.

```{r}
# your code
```

- Use the data frame `fem` to make a scatterplot of `height` and `mass`.

```{r}
# your code
ggplot(data = fem, aes(x = height, y = mass)) +
  geom_point()
```

- Find out how to make another scatterplot of `height` and `mass`, 
 using `geom_text()` to display the names of the individuals

```{r}
# your code
ggplot(data = fem, aes(x = height, y = mass)) +
  geom_point() + geom_text(aes(label = name))
```

- Get a scatter plot of `height` and `mass`, for ALL the females, displaying
their names with `geom_label()`.

```{r}
# your code
ggplot(data = filter(starwars, sex == 'female'), aes(x = height, y = mass)) +
  geom_point() + geom_label(aes(label = name))
```

- Get a histogram of `mass` (for all individuals).

```{r}
# your code
ggplot(data = starwars, aes(x = mass)) + geom_histogram(binwidth = 60)
```

- Get a density plot of `height` (for all individuals).

```{r}
# your code
ggplot(starwars, aes(x = height)) + geom_density(kernel = "gaussian")
```

- Get a barchart of the `gender` frequencies (for all individuals).

```{r}
# your code
ggplot(starwars, aes(x = gender)) + geom_bar()
```

