---
title: "Lab 4: More Data Tables"
subtitle: "Stat 133, Fall 2020"
output: 
  github_document:
    keep_html: true
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```

> ### Learning Objectives
>
> - How to import data tables in R
> - Get to know base reading tables functions
> - Pipes with `"dplyr"`


### General Instructions

- Write your descriptions, explanations, and code in an `Rmd` (R markdown) file.
- Name this file as `lab04-first-last.Rmd`, where `first` and `last` are your
first and last names (e.g. `lab04-gaston-sanchez.Rmd`).
- Knit your `Rmd` file as an html document (default option).
- Submit your `Rmd` and `html` files to bCourses, in the corresponding lab assignment.



## 1) Importing Data Tables

Data sets come in many different presentations. One common format is that of a 
data table---like a spreadsheet---although you can find data in other formats 
(especially when the data is in its raw version). However, once you get to the 
analysis stage, you will likely be dealing with some sort of tabular format.

Because data tables are so ubiquituous, it is fundamental that you learn how to
import them in R. In the first part of today's lab, we are going to review 
various aspects that have to do with reading in tables in R.

We will talk about the following functions (and packages)

- `read.table()`
- `read.csv()`
- `read.delim()`
- `read.fwf()`
- Optional: R package `"readr"`



## 2) Abalone Data Set

The first data set to consider is the __Abalone Data Set__ that is part of the
[UCI Machine Learning Repository](http://archive.ics.uci.edu/ml/datasets/Abalone)

The location of the data file is:

[http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data](http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data)

The location of the data dictionary (description of the data) is:

[http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.names](http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.names)

Use a text editor (e.g. the one in Rstudio) to look at both the dataset file `abalone.data`, and the file with its description `abalone.names`, and answer 
the following questions:

- What's the character delimiter?
- Is there a row for column names?
- Are there any missing values? If yes, how are they encoded?
- What are the data types of each column?


### 2.1) Getting a Local Copy of the Data

One basic way to read this file in R, is by passing the url location of the 
file directly to any of the `read.table()` functions:

```{r}
url <- "http://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data"
abalone <- read.table(url, sep = ",")
```


My suggestion when reading datasets from the Web, is to always try to get a local
copy of the data file in your machine (as long as you have enough free space
to save it in your computer). To do this, you can use the function `download.file()`
and specify the url address, and the name of the file that will be created in
your computer. For instance, to save the abalone data file in __your working 
directory__, type the following commands on the console:

Now that you have a local copy of the dataset, you can read it with 
`read.table()` like so:

```{r}
# reading data from your working directory
abalone <- read.table("abalone.data", sep = ",")
```


Once you read a data table, you may want to start looking at its contents,
usually taking a peek at a few rows. This can be done with `head()` and/or 
with `tail()`:

```{r}
head(abalone)
```


```{r}
tail(abalone)
```


Likewise, you may also want to examine how R has decided to take care of the
storage details (what data type is used for each column?). Use the function
`str()` to check the structure of the data frame:

```{r}
# check data frame's structure
str(abalone, vec.len = 1)
```



### 2.2) Detailed information about the columns 

So far we have been able to read the data file in R. But we are missing a few 
things. First, we don't have names for the columns. Second, it would be nice
if we could specify the data types of each column instead of letting R guess
how to handle each data type.

According to the description of the Abalone data set, the columns represent 
these variables:

| Name           | Data Type  |
|----------------|------------|
| Sex		     | nominal    |
| Length		 | continuous |
| Diameter	     | continuous |
| Height		 | continuous |
| Whole weight	 | continuous |
| Shucked weight | continuous |
| Viscera weight | continuous |
| Shell weight   | continuous |
| Rings		     | integer    |


### 2.3) Your Turn

- Create a vector `column_names` for the name of each column. Use the variable names displayed in the section "7. Attributes Information":

    + `sex`
    + `length`
    + `diameter`
    + `height`
    + `whole`
    + `shucked`
    + `viscera`
    + `shell`
    + `rings`

```{r}
# your code
column_names <- c('sex', 'length', 'diameter', 'height', 'whole', 'shucked',
                  'viscera', 'shell', 'rings')
column_names
```


- Create another vector `column_types` with R data types (e.g. `character`, `real`, `integer`). Match the R data types with the suggested type in "7. Attributes Information"  (nominal = `character`, continuous = `real`, integer = `integer`).

```{r}
# your code
column_types <- c(sex = 'character', length = 'real', diameter = 'real', 
                 height = 'real', whole = 'real', shucked = 'real',
                 viscera = 'real', shell = 'real', rings = 'integer')
column_types
```

- Optionally, you could also specify a type "factor" for the variable `sex` since
this is supposed to be in nominal scale (i.e. it is a categorical variable). 
Also note that the variable `rings` is supposed to be integers, therefore we 
can choose an `integer` vector for this column.

```{r}
# your code
sex_factor <- c(sex = 'factor', length = 'real', diameter = 'real', 
                 height = 'real', whole = 'real', shucked = 'real',
                 viscera = 'real', shell = 'real', rings = 'integer')
sex_factor
```


- Look at the documentation of the function `read.table()` and try to read the `abalone.data` table in R. Find out which arguments you need to specify so that you pass your vectors `column_names` and `column_types` to `read.table()`. Read in the data as `abalone`, and then check its structure with `str()`.

```{r}
# your code
abalone  <- read.table(file = 'abalone.data', header = TRUE, sep = ',',
                       col.names = column_names, colClasses = column_types)
str(abalone, vec.len = 1)
```


- Now re-read `abalone.data` with the `read.csv()` function. Name this data as `abalone2`, and check its structure with `str()`.

```{r}
# your code
abalone2 <- read.csv('abalone.data', header = TRUE)

str(abalone2, vec.len = 1)
```


- Look at the documentation of the function `read.table()` and find how to read just the first 10 lines in `abalone.data`? Name this data as 
`abalone10`, and check its structure with `str()`.

```{r}
# your code
abalone10 <- read.table(file = 'abalone.data', header = TRUE, sep = ',',
                       col.names = column_names, colClasses = column_types,
                       nrows = 10)
abalone10

str(abalone10, vec.len = 1)
```


- Look at the documentation of the function `read.table()` and find how to skip the first 10 lines in `abalone.data`, in order to read the next 10 lines (lines 11-20)? Name this data as `abalone20`, and check its structure with `str()`.

```{r}
# your code
abalone20 <- read.table(file = 'abalone.data', header = TRUE, sep = ',',
                       col.names = column_names, colClasses = column_types,
                       skip = 10, nrows = 10)
abalone20

str(abalone20, vec.len = 1)
```


- Look at the documentation of the function `read.table()` and find how to 
import  just the first four columns of `abalone.data`. Name this data as `aba4`,
and check its structure with `str()`.

```{r}
# your code
aba4 <- read.table(file = 'abalone.data', header = TRUE, sep = ',',
                   col.names = column_names, 
                   colClasses = c(rep("character", 1), 
                                  rep("double", 3),
                                  rep("NULL", 5)))

str(aba4, vec.len = 1)
```


- How would you import the first, the third, and the fifth columns of 
`abalone.data`. Name this data as `aba3`, and check its structure with `str()`.

```{r}
# your code
aba3 <- read.table(file = 'abalone.data', header = TRUE, sep = ',',
                   col.names = column_names, 
                   colClasses = c(rep("character", 1), 
                                  rep("NULL", 1), rep("double", 1),
                                  rep("NULL", 1), rep("double", 1),
                                  rep("NULL", 4)))

str(aba3, vec.len = 1)
```


- Use `summary()` on `abalone` to get summary statistics of the columns

Your output does not have to be in the 
same format of the table below. The important thing is that you begin learning 
how to manipulate columns (or vectors) of a data.frame.

```
       Length Diam  Height  Whole  Shucked  Viscera    Shell    Rings
Min	   0.075  0.055	0.000   0.002	 0.001    0.001	   0.002	    1
Max	   0.815  0.650	1.130   2.826	 1.488    0.760	   1.005	   29
Mean   0.524  0.408	0.140   0.829	 0.359    0.181	   0.239	9.934
```

```{r}
# your code
abalone %>% summary()
```

- Use the dollar operator `$` to select at least three columns, 
and pass them to functions such as `min()`, `max()`, etc, to compute 
descriptive statistics

```
       length diam  height  whole  shucked  viscera  shell  rings
Min	   0.075  0.055	0.000   0.002	 0.001    0.001	   0.002	    1
Max	   0.815  0.650	1.130   2.826	 1.488    0.760	   1.005	   29
Mean   0.524  0.408	0.140   0.829	 0.359    0.181	   0.239	9.934
SD	   0.120  0.099	0.042   0.490	 0.222    0.110	   0.139	3.224
```

```{r}
# your code
min(abalone$length)
max(abalone$height)
mean(abalone$shucked)
```


- Use `summary()` and dollar operator `$` to select columns sex, length, and
diameter from `abalone`, and obtain summary statistics.

```{r}
# your code
summary(abalone$sex)
summary(abalone$length)
summary(abalone$diameter)
```

- Use `hist()` to plot histograms of length, and diameter (use dollar operator 
`$` to select the corresponding column).

```{r}
# your code
hist(x = abalone$length)

hist(x = abalone$diameter)
```




## 3) More `"dplyr"`

This part of the lab involves working the pipe operator `%>%` which 
allows you write function calls in a more human-readable way. This becomes
extremely useful in `"dplyr"` operations that require many steps.

The behavior of `"dplyr"` is functional in the sense that function calls don't 
have side-effects. This implies that you must always save their results in order 
to keep them in an object (in memory). 

The "ugly" side of this functional behavior is that it doesn't lead to 
particularly elegant code, especially if you want to do many operations at once. 
You either have to do it step-by-step:

```{r echo = FALSE, eval = FALSE}
# manipulation step-by-step
dat1 <- group_by(abalone, sex)
dat2 <- select(dat1, sex, length, diameter, height)
dat3 <- summarise(dat2,
  avg_length = mean(length, na.rm = TRUE),
  avg_diam = mean(diameter, na.rm = TRUE),
  avg_height = mean(height, na.rm = TRUE))
dat4 <- arrange(dat3, avg_height)
dat4
```

Or if you don't want to name the intermediate results, you need to wrap the 
function calls inside each other:

```{r echo = FALSE, eval = FALSE}
# inside-out style (hard to read)
arrange(
  summarise(
    select(
      group_by(abalone, sex),
      sex, length, diameter, height
    ),
    avg_length = mean(length, na.rm = TRUE),
    avg_diam = mean(diameter, na.rm = TRUE),
    avg_height = mean(height, na.rm = TRUE)
  ),
  avg_height
)
```

This is difficult to read because the order of the operations is from inside 
to out. Thus, the arguments are a long way away from the function. 
To get around this problem, `"dplyr"` provides the `%>%` operator from 
`"magrittr"`.

`x %>% f(y)` turns into `f(x, y)` so you can use it to rewrite multiple 
operations that you can read left-to-right, top-to-bottom:

```{r echo = FALSE, eval = FALSE}
# using %>%
abalone %>% 
  group_by(sex) %>%
  select(sex, length, diameter, height) %>%
  summarise(
    avg_length = mean(length, na.rm = TRUE),
    avg_diam = mean(diameter, na.rm = TRUE),
    avg_height = mean(height, na.rm = TRUE)) %>%
  arrange(avg_height)
```


### 3.1) Your Turn: 

Use `"dplyr"` functions and the piper operator `"%>%"` to perform the following 
operations:

- slice the data to get the first 5 rows

```{r}
# your code
abalone %>% slice(1:5)
```

- slice the data to get the first 5 rows, arranged by height

```{r}
# your code
abalone %>% slice(1:5) %>% arrange(height)
```

- create a data frame `male_female` by filtering the data with Male and Female abalones, and display its dimensions with `dim()`

```{r}
# your code
male_female <- abalone %>% filter(sex == 'M' | sex == 'F')
dim(male_female)
```

- create a data frame `rings1` by filtering the data with those abalones having 25 Rings or more, containing columns about their sex, and rings (in that order).

```{r}
# your code
rings1 <- abalone %>% filter(rings >= 25) %>% select('sex', 'rings')
rings1
```

- create a data frame `rings2` by filter the data with those infant abalones with more than 3 Rings and less than 6, containing columns sex, rings, and diameter,
with rows arranged by diameter in increasing order.

```{r}
# your code
rings2 <- abalone %>% filter(rings > 3 & rings < 6) %>% 
  select('sex', 'rings', 'diameter') %>% arrange(diameter)
```


- Compute the median length, and median height, grouped by number of rings, of infant abalones only.

```{r}
# your code
abalone %>% filter(rings > 3 & rings < 6) %>% group_by(rings) %>%
  summarize(median_length = median(length), median_height = median(height))
```




## 4) Optional: Pittsburgh Bridges Data Set 

This part of the lab is provided as an optional challenge to give you more practice importing data tables, and manipulating data frames with `"dplyr"` and the piper operator.

The data set is part of the UCI Machine Learning Repository:

[http://archive.ics.uci.edu/ml/datasets/Pittsburgh+Bridges](http://archive.ics.uci.edu/ml/datasets/Pittsburgh+Bridges)

The data Description is here:

[http://archive.ics.uci.edu/ml/machine-learning-databases/bridges/bridges.names](http://archive.ics.uci.edu/ml/machine-learning-databases/bridges/bridges.names)

The Data file is here:

[http://archive.ics.uci.edu/ml/machine-learning-databases/bridges/bridges.data.version1](http://archive.ics.uci.edu/ml/machine-learning-databases/bridges/bridges.data.version1)


### 4.1) Your Turn

Read the description, and take a look at the data set:

- Are there column names?

- What is the field separator?

- Are there any missing values?

- What is the character for missing values (if any)?

- What is the data type of each variable (i.e. column)?

- Download a copy of the data to your computer (use `download.file()`)
and save it in a file named `bridges.data.version1`



### 4.2) Reading the Data

- Create a vector of column names

```{r}
# your code
```

- Create a vector of column types

```{r}
# your code
```

- Use the function `read.table()` to import the data. Name it `bridges`.

```{r}
# your code
```



### 4.3) Research Questions

Use `"dplyr"` functions and the piper operator to find:

- Year of the oldest bridge

```{r}
# your code
```

- Year of the most recent erected bridge

```{r}
# your code
```

- Frequency of bridges by purpose

```{r}
# your code
```

- Frequency of materials

```{r}
# your code
```

- Average length of the bridges

```{r}
# your code
```

- Plot a timeline: year -vs- length

```{r}
# your code
```
