---
title: "Lab 8: Functions and Loops"
subtitle: "Stat 133, Fall 2020"
output: html_document
urlcolor: blue
---

> ### Learning Objectives
>
> - Get familiar with the syntax of a `for` loop
> - Get familiar with the syntax of a `while` loop
> - Get familiar with the syntax of a `repeat` loop


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### General Instructions

- Write your descriptions, explanations, and code in an `Rmd` (R markdown) file.
- Name this file as `lab08-first-last.Rmd`, where `first` and `last` are your
first and last names (e.g. `lab08-gaston-sanchez.Rmd`).
- Knit your `Rmd` file as an html document (default option).
- Submit your `Rmd` and `html` files to bCourses, in the corresponding lab assignment.



## About Loops

- Many times we need to perform a procedure several times
- In other words, we have to perform the same operation several times as long as some condition is fulfilled
- For this purpose we use loops
- The main idea is that of __iteration__ or __repetition__
- R provides three basic paradigms to handle this situations: `for`, `repeat`, `while`




### 1) Your Turn: Summation Series

Write a for loop to compute the following two series. Your loop should start at step $k=0$ and stop at step $n$. Test your code with
different values for $n$. And store each k-th term at each iteration.
Does the series converge as $n$ increase?

#### 1.a) summation series

$$
\sum_{k=0}^{n} \frac{1}{2^k} = 1 + \frac{1}{2} + \frac{1}{4} + \frac{1}{8} + \dots + \frac{1}{2^n}
$$
```{r}
n <- 3 #example value for n
sum <- 0
k <- 0
while (k <= n){
  sum <- sum + 1/(2**k)
  k <- k + 1
}
print(sum)

```
The series converges as n increases.


#### 1.b) summation series


$$
\sum_{k=0}^{n} \frac{1}{9^k} =1 + \frac{1}{9} + \frac{1}{81} + \dots + \frac{1}{9^n}
$$

```{r}
n <- 3 #example value for n
sum <- 0
k <- 0
while (k <= n){
  sum <- sum + 1/(9**k)
  k <- k + 1
}
print(sum)
```
The series converges as n increases.

### 2) Your Turn: Arithmetic Series

Write a for loop to compute the following arithmetic series $a_n = a_1 + (n-1)d$ when $a_1 = 3$, and $d = 3$. For instance: 3 + 6 + 9 + 12 + 15 + ...

$$
a_n = a_1 + (n-1)d
$$

 
Test your code with different values for $n$. And store each n-th term at each iteration.
Does the series converge as $n$ increase?

```{r}
n <- 4 #example value for n
arith_series <- 0
a1 <- 3
d <- 3
for (pos in 1:n){
  arith_series <- arith_series + a1 + (n - 1)*d
  n <- n - 1
}
print(arith_series)
```
The series does not converge as n increases.

### 3) Your Turn: Geometric Sequence

A sequence such as $3, 6, 12, 24, 48$ is an example of a geometric sequence. 
In this type of sequence, the $n$-th term is obtained as: 

$$
a_n = a_1 \times r^{n-1}
$$

where: $a_1$ is the first term, $r$ is the common ratio, and $n$ is the number of 
terms.

Write a for loop to compute the sum of the first $n$ terms of:
3 + 6 + 12 + 24 + ... 
Test your code with different values for $n$. Does the series converge as $n$ 
increase?

```{r}
n <- 3 #example value for n
geom_seq <- 0
a1 <- 3
ratio <- 6/3
for (pos in 1:n){
  geom_seq <- geom_seq + a1*(ratio**(n-1))
  n <- n - 1
}
print(geom_seq)
```
The series does not converge as n increases.


### 4) Your Turn: Sine Approximation

Consider the following series that is used to approximate the function $sin(x)$:

$$
sin(x) \approx x - \frac{x^3}{3!} + \frac{x^5}{5!} - \frac{x^7}{7!} + \dots
$$

Write a `for` loop  to approximate $sin(x)$. Try different number of terms, 
$n = 5, 10, 50, 100$. Compare your results with those of `sin()` function.

```{r}
x <- 5 #example value for x
sin_func <- 0
degree <- 1
degree2 <- 3 
for (pos in 1:x){
  sin_func <- sin_func + ((x**degree)/factorial(degree) - (x**degree2)/factorial(degree2)) 
  degree <- degree + 4
  degree2 <- degree2 + 4
}
print(sin_func)

sin(5)
```


### 5) Your Turn: Average

The average of $n$ numbers $x_1, x_2, \dots, x_n$ is given by the following 
formula:

$$
\bar{x} = \frac{1}{n} \sum_{i=1}^{n} x_i = \frac{x_1 + x_2 + \dots + x_n}{n}
$$

Write R code, using each type of loop (e.g. `for`, `while`, `repeat`) to 
implement the arithmetic mean of the vector `x = 1:100`

```{r}
#for loop
sum <- 0
for (pos in 1:100){
  sum <- sum + pos
  avg <- sum/100
}
print(avg)
```
```{r}
#while loop
sum <- 0
x = 1
while (x < 101){
  sum <- sum + x
  avg <- sum/100
  x <- x + 1
}
print(avg)
```

```{r}
#repeat loop
sum <- 0
x = 1
repeat{
  sum <- sum + x
  avg <- sum/100
  x <- x + 1
  if (x > 100) {
    break
  }
}
print(avg)
```


### 6) Your Turn: Standard Deviation

The sample standard deviation of a list of $n$ numbers $x_1, x_2, \dots, x_n$ 
is given by the following formula:

$$
SD = \sqrt{ \frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})^2 }
$$

Write R code, using each type of loop (e.g. `for`, `while`, `repeat`) to 
implement the sample standard deviation of the vector `x = 1:100`

```{r}
#for loop
avg <- 50.5 #average of the vector 1 to 100
sum_diff <- 0 
for (pos in 1:100){
  sum_diff <- sum_diff + (pos - avg)**2
  sd <- ((1/(100-1))*sum_diff)**0.5
}
print(sd)
```
```{r}
#while loop
avg <- 50.5 #average of the vector 1 to 100
sum_diff <- 0
x = 1
while (x <= 100){
  sum_diff <- sum_diff + (x - avg)**2
  sd <- ((1/(100-1))*sum_diff)**0.5
  x <- x + 1
}
print(sd)
```
```{r}
#repeat loop
avg <- 50.5 #average of the vector 1 to 100
sum_diff <- 0
x = 1
repeat{
  sum_diff <- sum_diff + (x - avg)**2
  x <- x + 1
  sd <- ((1/(100-1))*sum_diff)**0.5
  if (x > 100) {
    break
  }
}
print(sd)
```


### 7) Your Turn: Geometric Mean

The geometric mean of $n$ numbers $x_1, x_2, \dots, x_n$ is given by the 
following formula:

$$
\bar{x} = \left ( \prod_{i=1}^{n} x_i \right )^{1/n}
$$

Write R code, using each type of loop (e.g. `for`, `while`, `repeat`) to 
implement the geometric mean of the vector `x = 1:50`

```{r}
#for loop
product <- 1
for (pos in 1:50){
  product <- product * pos
  geom_mean <- product**(1/pos)
}
print(geom_mean)
```

```{r}
#while loop
product <- 1
x = 1
while (x <= 50){
  product <- product * x
  geom_mean <- product**(1/x)
  x <- x + 1
}
print(geom_mean)
```

```{r}
#repeat loop
product <- 1
x = 1
repeat{
  product <- product * x
  geom_mean <- product**(1/x)
  x <- x + 1
  if (x > 50) {
    break
  }
}
print(geom_mean)
```

