---
title: "Lab 10: Regular Expressions 1"
subtitle: "Stat 133, Spring 2021"
output: 
  github_document:
    keep_html: true
urlcolor: blue
---

> ### Learning Objectives:
>
> - Get started with `stringi` and `stringr`
> - Get to know the basics of Regex :
> - `Anchors`
> - `Character classes`
> - `Quantifiers`
> - `Symbols`
> - Use regex with dataframes


### General Instructions

- Write your descriptions and code in an `Rmd` (R markdown) file.
- Name this file as `lab10-first-last.Rmd`, where `first` and `last` are your
first and last names (e.g. `lab10-gaston-sanchez.Rmd`).
- Knit your `Rmd` file as an html document (default option).
- Submit your `Rmd` and `html` files to bCourses, in the corresponding lab assignment.


## 1) Introduction

**Regular expression** also known as **Regex** is a tool for describing patterns in strings. The package associated with regex in R are stringr and stringi. Regex is also commonly used in other languages with perhaps slight syntactic changes.

But before we introduce you to regex, let's work with string functions in R.


### 1.1) stringi and stringr functions

> `str_detect()` helps detect string patterns in a string. Ensure you have installed stringi and stringr packages. 

```{r}
library(stringr)

city = "Berkeley"
str_detect(city, "kele")
```

We could pass a vector of cities as an arguement.

```{r}
city = c("Berkeley", "New York City", "Palo Alto", "New Delhi")
str_detect(city, "New")
```

> str_extract() helps extract a substring from a string and str_locate() helps with the start and end points of a pattern.

```{r}
city = "Berkeley"
str_extract(city, "ley")
str_locate(city, 'ley')
```

```{r}
city = c("Berkeley", "New York City", "Palo Alto", "New Delhi")
str_extract(city, "el")
```

> str_sub() returns a substring of a string given start and end position.

```{r}
city = c("Berkeley", "New York City", "Palo Alto", "New Delhi")
str_sub(city, start = 6, end = 8)
```


### 1.2) Your turn

#### 1.2.a) 

Print only those cities that start with `"New"` from `city`. Hint: Using `str_detect()`, we could create a mask (a vector of `TRUE` and `FALSE`) that could then be subsetted to get the city names starting with `"New"`.

```{r}
# your code
city = c("Berkeley", "New York City", "Palo Alto", "New Delhi")
city[str_detect(city, "New")]
```

#### 1.2.b)

Using `str_match()` explained in cheatsheet, check if `city` has any matches for `Berkeley`.

```{r}
# your code
city = c("Berkeley", "New York City", "Palo Alto", "New Delhi")
str_match(city, "Berkeley")
```



## 2) Using regex

Consider the question above where we printed cities starting with `"New"`. Using regex, we could do it more directly.

To let R know that we are passing regex instead of a substring, we use `regex()`.

```{r}
str_extract(city, regex("el"))
```

```{r}
str_extract(city, regex("^New [a-zA-Z]*"))
```

> In the second example, `^` symbol is an anchor that matches any string starting with `"New"`. `[a-zA-Z]` is a set of characters that could appear and `*` is the quantifier, which specifies how many of those characters in square bracket appear.


### 2.1) Anchors

Anchors help us assert the position, i.e., the beginning or end of the string.

|Anchor |Description| Example| 
|----------|---------|---------|
| `^` | Matches any string starting with a substring | `^New` |
| `$` | Matches any string ending with a substring | `y$` |
|Exact Match `^$` | Matches the string that starts and ends with substring | `^Hi There$` |


```{r}
banks = c("Bank of America", "Bank of the West", "Citibank", 
          "TD Bank", "Bank of England", "People\'s United Bank")

#banks starting with Bank
str_detect(banks, regex("^Bank"))

#banks ending with Bank (note: citibank is False as b is lowercase)
str_detect(banks, regex("Bank$"))

#bank called "TD Bank"
str_detect(banks, regex("^TD Bank$"))

str_detect(banks, "TD Bank")
```


### 2.2) Character Classes

Character classes match any characters given a class

| Character Class | Match |
|-----------------|-------|
| `[aeiou]` | Matches vowels |
| `[0-9]` | Matches digits |
| `[a-z]` | Matches lower case letters |
| `[A-Z]` | Matches upper case letters |
| `[a-zA-Z]` | Matches both lower and upper case letters |
| `[a-g]` | Matches characters a through g | 

In this example, we will try to print the names of banks starting with `"Bank"`

```{r}
#Note the whitespace after Z in the regular expression
str_extract(banks, regex("^Bank[a-zA-Z ]*"))

#Without the whitespace after Z, it will not give us the desired output
str_extract(banks, regex("^Bank[a-zA-Z]*"))
```

> In the previous example, `*` acted as a quantifier. It means any number of the group within square brackets could appear after `"Bank"`.  Refer next section on quantifiers. 


### 2.3) Quantifiers

Quantifiers quantify the number of instances of a character, group or character class. Your quantifier should be placed after the character/group/character class that is being quantified.

|Quantifier|Description|
|----------|----------|
| `P*` | 0 or more instances of P|
| `P+` | 1 or more instances of P|
| `P?` | 0 or 1 instance of P|
| `P{m}` | Exactly m instances of P|
| `P{m,}` | At least m instances of P|
| `P{m,n}` | Between m and n instances of P|

Let us try to print bank names that have less than 10 characters in them. 

```{r}
str_extract(banks,regex("^[a-zA-Z ]{1,9}$"))
```

> Notice that white space is included after `A-Z`
> Also note that we need to use anchors `^ $`, otherwise the first 10 characters of all banks will be printed.


### 2.4) Logical Operators

| Operator | Usage | Example | Explanation |
|----------|-------|---------|-------------|
| `^`  | Not | `[^A-Za-z0-9]` | Identifies characters that are not alphanumeric |
|` \|` | Or  | `(Apple\|Orange)` | Identifies cases that have Apple or Oranges or both |

There is no AND operation in regex.

```{r}
print(banks)

#Detect all bank names containing Bank or bank
str_detect(banks, regex("Bank|bank"))

#Detect all bank names containg special characters excluding whitespace
str_detect(banks, regex("[^A-z0-9 ]"))

```



## 3) More examples

```{r}
# cities ending with y
str_detect(city, regex("y$"))

#List the names of first 10 nba players 
# with college name starting with "University"
library(dplyr)
nba = read.csv("C:/Users/felic/nba2018-players.csv")

nba %>% 
  filter(str_detect(college, regex("^University"))) %>% 
  select(player) %>% 
  head(10)

# List the names of colleges of nba players containing 
# special characters (don't consider whitespace and , 
# as special character here)
nba %>% 
  filter(str_detect(college, regex("[^A-Za-z, ]"))) %>% 
  select(college)
```

Let us try to print bank names with 3 words in them. We create groups for each word and make them repeat 3 times. 

> The first group is ([a-zA-Z']* ){2}. Here {2} refers to the number of times the group is repeated. [a-zA-Z']* means characters and ' can appear any number of times. A whitespace is included because after the first two words, a whitespace appears. 
> the second group is ([a-zA-Z]*){1}. We repeat this group only once. Note that we don't have whitespace in this group.

```{r}
str_extract(banks, regex("^([a-zA-Z']* ){2}([a-zA-Z]*){1}$"))
```


### 3.1) Your turn

#### 3.1.a)

Using `nba` dataframe, print names of Players containing `Marcus`

```{r}
# your code
nba %>% 
  filter(str_detect(player, regex("Marcus"))) %>% 
  select(player)
```

#### 3.1.b)

Using `nba` dataframe, print names of Universities that contain `California` or `Los Angeles` in them.

```{r}
# your code
nba %>% 
  filter(str_detect(college, regex("California|Los Angeles"))) %>% 
  select(college) 
```

#### 3.1.c)

Consider the variable below `myVar`. Using str_detect and regular expression, detect those strings that contain at least one `z` and a maximum of three `z`. Your output should be `TRUE TRUE TRUE FALSE`

```{r}
myVar = c("bizarre", "bizzarre", "bizzzarrre", "bizzzzare")
```

```{r}
# your code
str_detect(myVar, regex("[z]{1,3}[^z]{4,}"))
str_detect(myVar, regex("biz{1,3}a"))
```

#### 3.1.d)

Consider the output of ls statement from command line. We have stored it in `file_names`, a vector containing file names. Detect file names that start with `STAT154` and have extension `.csv`
\
Expected Output: `TRUE FALSE FALSE TRUE FALSE TRUE`

```{r}
file_names = c(
  "STAT154nba.csv", 
  "test0102_STAT154.csv", 
  "STAT154myfile.csv.tmp", 
  "STAT154_lab.csv", 
  "STAT154_HW.pdf", 
  "STAT154_test0102.csv")
```

```{r}
# your code
str_detect(file_names,
           regex("^([STAT154]){1,}([a-zA-Z0-9|[:punct:]]*)[.csv]$"))
```

#### 3.1.e)

For the variable `newVar`, identify strings that start with a number. Your output should be `TRUE FALSE TRUE TRUE FALSE`. 

```{r}
newVar = c(
  "1 Student(s)", 
  "None but 1 Student(s)", 
  "5 Students", 
  "120! Students", 
  "Two Students")
```

```{r}
# your code
str_detect(newVar, regex("^[0-9]"))
```

#### 3.1.f)

Detect phone numbers from a given list. Note that these won't have international codes. Your output for given test case should be `TRUE TRUE TRUE TRUE FALSE FALSE`
Hint: The last string in the vector has more than 10 digits.

```{r}
phone_num = c(
  "401-501-1111", 
  "(401)501-1111", 
  "401 501 1111", 
  "4015011111", 
  "+408-501-1111", 
  "40850211111")
```

```{r}
# your code
str_detect(phone_num, regex("^([1-9|\\(]){1}[0-9{9}|[:punct:]{2,3}]"))
```

#### 3.1.g)

In the solution you gave above, can you think of cases where your expression would fail to detect an incorrect phone number? If yes, how could you improve it?

```{r}
# No code, but some text
# Yes, one example would be that a phone number with the format (xxx)-xxx-xxxx would have 4 punctuations (2 parenthesis and 2 dashes) and therefore would not work with my current code. In order to improve it, I could expand the current quantifier {2,3} to include 4.
```

#### 3.1.h)

In column of a dataframe, we have stored strings that tell us if someone likes orange juice, apple juice etc. Extract the names of students who like orange juice or oranges. Expected Output: Annie Harry
Hint: Use `str_detect` before using `str_extract()`

```{r}
myVar = c(
  "Annie likes Orange juice", "Sonny likes Apple juice", 
  "Katy dislikes Orange juice", "Harry likes Oranges", 
  "Charlie likes Apple juice", "Margo like Orange Pie")

df = data.frame(myVar)

mask <- str_detect(df[, 1], regex(" likes Orange"))
#mask <- c(TRUE, FALSE, FALSE, TRUE, FALSE, FALSE)
```

```{r}
# your code
str_extract(df[mask, ], regex("[A-z]+"))
```



## 4) Using Apply 

Apply functions are useful when you want to apply a certain operation to all the rows of a list or dataframe. In the example below, we will only consider the case of a data frame. Note that this idea would be useful for your next lab. Here we apply `str_locate()` to locate area code from phone numbers. Using the output of `str_locate()`, we could then extract the area code.

```{r}
phone_num = c(
  "401-501-1111", 
  "(401)501-1111", 
  "401 501 1111", 
  "401-5011111", 
  "+408-501-1111")

code_pos = lapply(
  phone_num, 
  function(x) str_locate(x, pattern = regex('[0-9]{3}[- )]*')))

# Use substring() to extract the area code by sharing the start 
# and end position with the function as arguments.
code_pos = do.call(rbind, code_pos)
substring(phone_num, code_pos[,1], code_pos[,2]-1)
```


### 4.1) Your Turn

Using the idea above, from the `nba` dataframe, pull the last names of all 
NBA players. Find the frequencies of all the last names. _Hint:_ Use `str_locate()` and `substring()`. Eg: str_locate(nba$player, ...). Alternatively, you may choose to try this with lapply() but this is slightly tricky.

```{r}
# your code
location_last_name <- str_locate(nba$player, regex("(?<=\\s)[:alpha:]")) 
last_names <- str_sub(nba$player, location_last_name[ , 1])
frequencies <- table(last_names)
frequencies
```
