---
title: 'Lab 12: A little bit of web scraping'
author: "Felicia Liu"
date: "4/23/2021"
output: html_document
---

```{r}
library(xml2)
library(rvest)
library(stringr)
```
# 1) Extract Information from Google Scholar

## 1.1.1) HTML source code

## 1.1.2) Download and read the web page with `read_html()`

```{r}
Sys.sleep(15)

scholar <- "https://scholar.google.com/citations?user="
author <- "YA43PbsAAAAJ&hl=en&oi=ao"
author_url <- paste0(scholar, author)

download.file(author_url, "google_scholar.html")
```

```{r}
parent_data <- read_html('google_scholar.html')
table_data <- html_table(parent_data)
```

## 1.2) 

a) Extract the Google scholar ID of the author from the URL. It contains 12 characters.

```{r}
str_extract(parent_data, "(?<=user=)[A-z0-9]{12}")
```
b) Extract and store the summary table of the citations in the top right corner of the user profile page

```{r}
table_data[[1]]
```
c) Extract and store all the titles/name/journal of the papers in a vector `long_info`

```{r}
long_info <- c(table_data[[2]][[1]])
long_info
```
d) Extract and store number of the citations in a vector `citation`

```{r}
citation <- c(table_data[[2]][[2]])
citation
```
e) Extract and store the years of publication in a vector `year`

```{r}
year <- c(table_data[[2]][[3]])
year
```

## 1.3) Alternative way of reading the nodes: html_nodes()

```{r}
table_body <- parent_data %>% html_nodes(xpath = '//*[@id="gsc_a_b"]')
table_body
```
```{r}
table_row <- parent_data %>% html_nodes(xpath = '//*[@id="gsc_a_tr"]')
table_row
```
## 1.4) Your turn

a) Extract information from the web page using html_nodes() and store the information in a dataframe

```{r}
table_citations <- parent_data %>% html_nodes(xpath = '//*[@id="gsc_a_t"]')
table_citations
```

b) Among the 20 articles, give the title of the most cited article?

ggplot2, ggplot2: elegant graphics for data analysis

c) Among the 20 articles, in which years has the author published more than two articles?

2009, 2011, 2014
